---
title: "ESM244 Lab 5"
author: "Sara Orofino"
date: "2/6/2019"
output: html_document
---


a. Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tseries)
library(forecast)


```

b. Get data

US Residential energy consumption values (Jan 1973 - Oct 2017), Trillion BTU (US Energy Information Administration)

```{r, message=FALSE}

energy <- read_csv("energy.csv") 

```



c. Create 'ts' time series data using information in the res_total column from 'energy'

```{r}

#Create vector of observations for time series, using energy res_total data, which is monthly data (frequency = 12), and it starts in January 1973


res_ts <- ts(energy$res_total, frequency = 12, start = c(1973,1)) 

# plot ts for exploratory graphics 
plot(res_ts)

```


d. Decompose ts

```{r}
# Decompose the time series data - just using decompose() in base r

res_dc <- decompose(res_ts)

# View the breakdown of observed vs trend vs seasonal vs random 
plot(res_dc)


# If there are cylces outside of seasonality they would show up in the random graph of the plot
# Pay attention to the scales because it can show whether the seasonal impacts are important on the same scale as the observed data or the trend. 
```

e. Other cool ways to present ts data 

```{r}

# month plot on original ts data NOT decomposed - shows how values are changing within each month 

monthplot(res_ts) #shows a big increase in energy usage in summer months 



```


```{r}

#ggseason plot - plots the time series within each year over the 12 months

ggseasonplot(res_ts) 

ggseasonplot(res_ts, polar = TRUE) # use polar = TRUE to plot a cyclical graph 
```


f. Simple moving averages 

forecast::ma() to specify windows for moving averages 

```{r}


sma_res <- ma(res_ts, order = 5)
plot(sma_res)

# Change the window size... order = 12

sma_res2 <- ma(res_ts, order = 12)
plot(sma_res2)

# Good to plot both the original ts data and the moving average together (exploratory graph)

plot(res_ts)
lines(sma_res, col = "red")

```

g. ACF (Autocorrelation) 

```{r}
# plots the autocorrelation for the time series data 
res_acf <- acf(res_ts)

# can see the big peaks every 12 months - strong correlation with observations that were 12 months ago
# Seasonality is important

# more information on the data (in forecasts package) ACF and PACF (partial autocorrelation)
# can finalize just like a ggplot 

ggtsdisplay(res_ts)
```

i. Forecasting with Holt-Winters exponential smoothing

Note: Forecasting/smoothing will not improve the data you're working with...

```{r}

res_hw <- HoltWinters(res_ts)
res_hw

# Additive data; gives the alpha, beta, and gamma parameters 

plot(res_hw)

#The plot shows in red how the model looks compared to the actual data 
```

Now some future forecasting with Holt-Winters

```{r}
# forecasting for 5 years on a monthly timescale h = 60

res_forecast <- forecast(res_hw, h = 60)
plot(res_forecast) 

#blue lines are the predictions for forecasted time; light and dark grey are showing the 80% and 95% confidence intervals 
# can transform this to a dataframe and use to make a plot in ggplot 
```

Check distribution of residuals:

```{r}

# residuals will indicate normality - graph on histogram and qqplot

par(mfrow = c(1,2))
hist(res_forecast$residuals) 
qqnorm(res_forecast$residuals)

```


Autoregressive integrated moving average (ARIMA) using auto.arima for p,d,q


```{r}

# Auto arima will give you pdq for non-season component (first) and seasonal component (second if there is one)

res_pdq <- auto.arima(res_ts)
res_pdq

```

Non-seasonal pdq (1,0,2)  
Seasonal pdq (0,1,1)   


Fit the ARIMA model using those pdq inputs:

```{r}

# List the non-seasonal pdq first as just order = c(p,d,q) and the seasonal second as list(order = c(p,d,q))

res_arima <- arima(res_ts, order = c(1,0,2), seasonal = list(order = c(0,1,1)))

#check out the residuals (exploratory)

par(mfrow = c(1,2)) #Use this line of code to specify the two graphs be placed side by side
hist(res_arima$residuals)
qqnorm(res_arima$residuals)

#Overall looks normal, not horribly skewed in either direction 
```

Now forecast with ARIMA outputs:

```{r}
# time of forecast(h) is in months because the data was collected monthly 

forecast_res <- forecast(res_arima, h = 72)
plot(forecast_res)

# Notice the ARIMA looks pretty similar to the Holt-Winters; because the data is very seasonal and highly predictable so both have trends that look similar 

# Would generally be a good idea to show both the ARIMA and the Holt-Winters in publication 
```

Now let's try viewing this in ggplot:

```{r}
# remember you can make the listed values from forecasting a dataframe to plot in ggplot:

res_df <- data.frame(forecast_res)

#Point.Forecast - the forecasted data points, H80/L80 shows the high and low of the 80% Confidence Interval

month_seq <- seq(1,72) #creates a series of months from 1-72 to match the predictions that were forecasted 
res_df_2 <- data.frame(month_seq, res_df)

ggplot(res_df_2, aes(x=month_seq, y=Point.Forecast)) +
  geom_line()

# can use geom_ribbon for the confidence intervals and finalizing as necessary 
```












